---
title: "annotation"
author: "Javier Villegas"
date: "2025-12-03"
output: html_document
---

```{r setup, include=FALSE}
# Load libraries
library(Seurat)
library(tidyverse)
library(reticulate)
library(anndata)
```

# MapMyCells annotation

## 1. Generate input file for MapMyCells

```{r mapmycells_input}

# Filtered Seurat object generated in the previous preprocessing script
KA_seurat_filtered <- readRDS("Filtering/KA_seurat_filtering.rds")

# Python environment used by anndata
# Run these lines once if the environment has not already been created:
#
# version <- "3.9.12"
# install_python(version)
# virtualenv_create("my-environment", version = version)

use_virtualenv("my-environment")

# Extract raw counts and transpose to cells x genes
dataQCt <- Matrix::t(
  as.matrix(
    GetAssayData(
      KA_seurat_filtered,
      assay = "RNA",
      layer = "counts"
    )
  )
)

# Convert to AnnData
ad <- AnnData(
  X = dataQCt,
  obs = data.frame(
    group = rownames(dataQCt),
    row.names = rownames(dataQCt)
  ),
  var = data.frame(
    type = colnames(dataQCt),
    row.names = colnames(dataQCt)
  )
)

# Save compressed h5ad file
write_h5ad(
  ad,
  "IAKA_seurat_MapMyCells.h5ad",
  compression = "gzip"
)

# Check file size
print(
  paste(
    "Size in MB:",
    round(file.size("IAKA_seurat_MapMyCells.h5ad") / 2^20)
  )
)
```

## 2. Run MapMyCells

The generated `.h5ad` file was uploaded manually to **MapMyCells** using the
10x Whole Mouse Brain taxonomy. The resulting hierarchical mapping `.csv`
file was downloaded and used below.

## 3. Add MapMyCells annotations

```{r import_mapmycells}

# Read MapMyCells output
mapping <- read.csv(
  "MMC_output/IAKA_seurat_MapMyCells_HierarchicalMapping.csv",
  comment.char = "#"
)

# MapMyCells output should contain one row per nucleus
stopifnot(nrow(mapping) == ncol(KA_seurat_filtered))

# Match annotation rows to nuclei in the Seurat object
rownames(mapping) <- colnames(KA_seurat_filtered)

# Add MapMyCells annotation to Seurat metadata
KA_seurat_filtered <- AddMetaData(
  KA_seurat_filtered,
  metadata = mapping
)

# Check annotations
table(KA_seurat_filtered$class_name)
table(KA_seurat_filtered$subclass_name)

# Plot original MapMyCells annotations
DimPlot(
  KA_seurat_filtered,
  group.by = "class_name",
  label = TRUE,
  repel = TRUE
) + NoLegend()

DimPlot(
  KA_seurat_filtered,
  group.by = "subclass_name",
  label = TRUE,
  repel=TRUE
) + NoLegend()
```

# Rename MapMyCells annotations

The original MapMyCells subclasses were grouped into the cell classes used
for downstream analyses.

```{r rename_annotations}

# Rename / merge MapMyCells annotations
# Smaller MapMyCells subclasses were grouped into broader parent classes to obtain the cell subtype categories used for downstream analyses.

# Helper function to create a lookup table
make_lookup <- function(groups) {

  unlist(
    lapply(names(groups), function(group) {

      setNames(
        rep(group, length(groups[[group]])),
        groups[[group]]
      )

    })
  )
}


########################################################
# Neuronal classes

class_groups <- list(

  "CA1 Glut Neurons" = c(
    "016 CA1-ProS Glut"
  ),

  "CA2 Glut Neurons" = c(
    "025 CA2-FC-IG Glut"
  ),

  "CA3 Glut Neurons" = c(
    "017 CA3 Glut"
  ),

  "HPF CR Glut Neurons" = c(
    "036 HPF CR Glut"
  ),

  "SUB-ProS Glut Neurons" = c(
    "023 SUB-ProS Glut"
  ),

  "Dentate Gyrus Neurons" = c(
    "037 DG Glut"
  ),

  "DG-PIR Immature Neurons" = c(
    "038 DG-PIR Ex IMN"
  ),

  "IT-ET Glut Neurons" = c(
    "008 L2/3 IT ENT Glut",
    "019 L2/3 IT PPP Glut",
    "018 L2 IT PPP-APr Glut",
    "011 L2 IT ENT-po Glut",
    "003 L5/6 IT TPE-ENT Glut",
    "020 L2/3 IT RSP Glut",
    "007 L2/3 IT CTX Glut",
    "009 L2/3 IT PIR-ENTl Glut",
    "015 ENTmv-PA-COAp Glut",
    "006 L4/5 IT CTX Glut",
    "030 L6 CT CTX Glut",
    "022 L5 ET CTX Glut",
    "004 L6 IT CTX Glut",
    "024 L5 PPP Glut",
    "005 L5 IT CTX Glut",
    "021 L4 RSP-ACA Glut",
    "002 IT EP-CLA Glut",
    "012 MEA Slc17a7 Glut",
    "014 LA-BLA-BMA-PA Glut",
    "001 CLA-EPd-CTX Car3 Glut",
    "010 IT AON-TT-DP Glut",
    "013 COAp Grxcr2 Glut"
  ),

  "NP CT L6b Glut" = c(
    "028 L6b/CT ENT Glut",
    "029 L6b CTX Glut",
    "031 CT SUB Glut",
    "033 NP SUB Glut",
    "034 NP PPP Glut",
    "027 L6b EPd Glut",
    "032 L5 NP CTX Glut"
  ),

  "Sst GABA Neurons" = c(
    "056 Sst Chodl Gaba",
    "053 Sst Gaba"
  ),

  "Pvalb GABA Neurons" = c(
    "051 Pvalb chandelier Gaba",
    "052 Pvalb Gaba"
  ),

  "Lamp5 GABA Neurons" = c(
    "049 Lamp5 Gaba",
    "050 Lamp5 Lhx6 Gaba"
  ),

  "VIP GABA Neurons" = c(
    "046 Vip Gaba"
  ),

  "Immature GABA Neurons" = c(
    "045 OB-STR-CTX Inh IMN",
    "041 OB-in Frmd7 Gaba",
    "044 OB Dopa-Gaba",
    "039 OB Meis2 Thsd7b Gaba",
    "042 OB-out Frmd7 Gaba"
  ),

  "CTX-CGE GABA" = c(
    "048 RHP-COA Ndnf Gaba",
    "047 Sncg Gaba"
  ),


########################################################
# Non-neuronal classes

  "Oligo Progenitor Cells" = c(
    "326 OPC NN"
  ),

  "Oligodendrocytes" = c(
    "327 Oligo NN"
  ),

  "Astrocytes" = c(
    "318 Astro-NT NN",
    "319 Astro-TE NN"
  ),

  "Microglia" = c(
    "334 Microglia NN"
  ),

  "Vascular cells" = c(
    "333 Endo NN",
    "331 Peri NN",
    "329 ABC NN",
    "332 SMC NN",
    "335 BAM NN",
    "321 Astroependymal NN",
    "322 Tanycyte NN",
    "323 Ependymal NN",
    "338 Lymphoid NN",
    "337 DC NN",
    "330 VLMC NN"
  ),

  "Choroid Plexus" = c(
    "325 CHOR NN"
  )
)


# Convert MapMyCells subclass names to the classes used in the analysis
class_lookup <- make_lookup(class_groups)

KA_seurat_filtered$classes <- unname(
  class_lookup[KA_seurat_filtered$subclass_name]
)

# Subclasses not used in downstream analyses are labelled as Other
KA_seurat_filtered$classes[
  is.na(KA_seurat_filtered$classes)
] <- "Other"

KA_seurat_filtered$classes <- factor(
  KA_seurat_filtered$classes
)

# Check final annotation
table(KA_seurat_filtered$classes)

DimPlot(
  KA_seurat_filtered,
  group.by = "classes",
  label = TRUE,
  repel=TRUE
) + NoLegend
```

# Prepare final Seurat object

Subtypes represented by fewer than 10 nuclei were excluded from the
downstream analyses.

```{r final_object}

# Cell subtypes retained for downstream analysis
classes_filtering <- c(
  "Choroid Plexus",
  "Vascular cells",
  "Microglia",
  "Astrocytes",
  "Oligodendrocytes",
  "Oligo Progenitor Cells",
  "DG-PIR Immature Neurons",
  "Dentate Gyrus Neurons",
  "CTX-CGE GABA",
  "Immature GABA Neurons",
  "VIP GABA Neurons",
  "Lamp5 GABA Neurons",
  "Pvalb GABA Neurons",
  "Sst GABA Neurons",
  "NP CT L6b Glut",
  "IT-ET Glut Neurons",
  "SUB-ProS Glut Neurons",
  "HPF CR Glut Neurons",
  "CA3 Glut Neurons",
  "CA2 Glut Neurons",
  "CA1 Glut Neurons"
)

# Keep the cell classes used in the final analysis
KA_seurat_plots <- subset(
  KA_seurat_filtered,
  subset = classes %in% classes_filtering
)

# Remove unused factor levels
KA_seurat_plots$classes <- droplevels(
  KA_seurat_plots$classes
)

# Check number of nuclei in each retained class
table(KA_seurat_plots$classes)

# Optional: save object for downstream scripts
# saveRDS(KA_seurat_plots, "KA_seurat_plots.rds")
```